---
title: "Assessing Bay Area Housing and Income Equity and Demographics"
author: "Bella Raja"
output: html_document
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = F, warning= F, message=F, progress_bar =F)
```

```{r}
library(tidyverse)
library(censusapi)
library(tigris)
library(leaflet)
library(mapview)
library(sf)
library(tidycensus)
library(plotly)
library(dplyr)
library(tmap)
library(mapdeck)



Sys.setenv(CENSUS_KEY = "ffe5c482aad3c87fd918ba8a6bd5967ae4228456")
Sys.getenv("ffe5c482aad3c87fd918ba8a6bd5967ae4228456")
```



In this project, I wanted to analyze migration patterns in the Bay Area which led me to questions of housing and housing prices as related to income concentration. 
I also wanted to assess demographics in the Bay Area and examine different equity analyses and indices of segregation.
Finally, I related all of these contemporary findings to the legacy of historic redlining and economic risk "grades" given to different Bay Area regions. 


## Datasets for this project

For this project, I use various different types of census data as well as data from the 1937 Home Owners' Loan Corporation grading scheme of American neighborhoods.

Every year, the U.S. Census Bureau puts out and publishes the American Communities Survey (ACS) 1-year estimates for geographic areas with populations of 65,000 or more. This data is taken from random households and ask more detailed questions than the decennial census because its purpose is to track various economic and social changes rather than broad population counts. The ACS 5-year estimates are a bit more statistically reliable than the 1-year estimates because they gather data over consecutive years and also have larger sample sizes.

I use both American Communities Survey 1-year and 5-year data. In particular, 1-year estimates are helpful for understanding quickly changing geographic or demographic data, such as mobility or in-flow/out-flow data. 5-year estimates can provide interesting insights into broader trends in areas such as racial demographics or home values and annual income. Public Use Microdata Sample (PUMS) files are also included in these ACS reports. The PUMS data provides even more specific and granular data from individuals and households. This data is helpful for some of the more specific variables I measure, such as households with income less than 100K. 





## Bay Area Migration - 2019

```{r}
token <- "pk.eyJ1IjoiYmVsbGFyIiwiYSI6ImNrd3JjNmFkOTB2MmQycWp2Mm42OXc5dXgifQ.pttZ3IYhuP9cq19jSqvIzA"

bay_inflow <- get_flows(
  geography = "county",
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  geometry = TRUE,
  year = 2019
) %>%
  filter(variable == "MOVEDIN") %>%
  na.omit() %>%
  arrange(desc(estimate))

# finding inflows again but this time using tidycensus flows function
```





```{r}
bay_outflow <- get_flows(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  geometry = TRUE,
  key = "ffe5c482aad3c87fd918ba8a6bd5967ae4228456"
) %>%
  filter(variable == "MOVEDOUT") %>%
  na.omit() %>%
  arrange(desc(estimate))
view(bay_outflow)
```
These 3D Maps show the inflow and outflow for the Bay Area. 
The arcs show the top 50 inflow to and outflow from the Bay Area locations.
To navigate these maps, double tap or scroll up with your mouse to zoom in and scroll down with your mouse to zoom out on particular spots. 

Zoom in on the Bay Area to get a closer look at the top 50 destinations and places of departure.

Bay Area Inflow Locations - 2019                     
```{r}
bay_inflow_map <- bay_inflow  %>%
  slice_max(estimate, n = 50) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = token) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight")
bay_inflow_map
```
This map uses the 2019 ACS 1 year estimates to show the inflow of people moving to the Bay Area from the top 50 inflow origin locations.


Bay Area Outflow Locations- 2019
```{r}
bay_outflow_map <- bay_outflow  %>%
  slice_max(estimate, n = 50) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = token) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight")

bay_outflow_map 
```
This map uses the same data as the previous map to show the top 50 outflow locations for people moving out of the Bay Area.

In comparing these two maps, a lot of the 2019 inflow and outflow locations are very similar such as moving to or from San Francisco. There is more geographic distance in the inflow destinations, with more people moving from big cities like New York and Chicago. Whereas, the outflow destinations are closer to the Bay Area, with people moving to places like cities in Southern California or Las Vegas in Nevada (where housing prices and taxes are much lower but still close enough to potentially work remotely or stay near family).


```{r, progress_bar = F }
acs_vars_2019_1yr <-
  listCensusMetadata(
    name = "2019/acs/acs1",
    type = "variables"
  )

bay_mobility_current_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07010)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)
```

```{r}
bay_mobility_current_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07010)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"mobility","temp","income"),
    sep = "!!"
  ) %>% 
  mutate(
    income = ifelse(
      temp == "No income",
      temp,
      income
    ),
    mobility = ifelse(
      mobility %in% c("Same house 1 year ago:", "Moved within same county:"),
      "Here since last year",
      "Inflow"
    )
  ) %>% 
  filter(!is.na(income)) %>% 
  group_by(mobility, income) %>% 
  summarize(estimate = sum(estimate))

#Mobility in Bay Area
```

```{r}
bay_mobility_lastyear_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07410)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)
#Inflow since last year
```

```{r}
bay_mobility_lastyear_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07410)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"mobility","temp","income"),
    sep = "!!"
  ) %>% 
  mutate(
    income = ifelse(
      temp == "No income",
      temp,
      income
    ),
    mobility = ifelse(
      mobility %in% c("Same house:", "Moved within same county:"),
      "Here since last year",
      "Outflow"
    )
  ) %>% 
  filter(!is.na(income)) %>% 
  group_by(mobility, income) %>% 
  summarize(estimate = sum(estimate))

view(bay_mobility_lastyear_19)
#outflow
```
 



```{r}
bay_mobility_current_18 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2018,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07010)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA, NA, "mobility","temp","income"),
    sep = "!!"
  ) %>% 
  mutate(
    income = ifelse(
      temp == "No income",
      temp,
      income
    ),
    mobility = "Here last year"
  ) %>% 
  filter(!is.na(income)) %>% 
  group_by(mobility, income) %>% 
  summarize(estimate = sum(estimate))
#total pop 2018
```

#Bay Area Flows by Income Tier 
```{r}
bay_flows_19 <-
  rbind(
    bay_mobility_current_18,
    bay_mobility_lastyear_19 %>% 
      filter(mobility == "Outflow"),
    bay_mobility_current_19 %>% 
      filter(mobility == "Inflow"),
    bay_mobility_current_19 %>% 
      group_by(income) %>% 
      summarize(estimate = sum(estimate)) %>% 
      mutate(mobility = "Here this year")
  ) %>% 
  pivot_wider(
    names_from = mobility,
    values_from = estimate
  ) %>% 
  mutate(
    `External net` = Inflow - Outflow,
    `Internal net` = `Here this year` - `Here last year` - `External net`,
  ) %>% 
  select(
    `Income tier` = income, 
    `Internal net`,
    `External net`,
    `Here last year`, 
    `Here this year`, 
    Outflow, 
    Inflow
  )


bay_flows_19
```

This table shows inflow and outflow across the Bay Area counties based on income tiers. This data comes from the ACS 1-year estimates of Geographic Mobility from 2018 to 2019. 
Those with a higher income have more inflow than outflow to the Bay Area, and those with lower incomes have more outflow than inflow to the Bay Area. With increasing housing costs and taxes, the Bay Area is a very expensive place to live. 



## Assesing Bay Area Income and Housing
```{r}
housing_value <- get_acs(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variables = "B25077_001",
  survey = "acs1",
  geometry = TRUE,
  resolution = "20m"
)
```
Median Home Values by Bay Area County - 2019
```{r}
housing_pal <- colorNumeric(
  palette = "plasma",
  domain = housing_value$estimate
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = housing_value,
              color = ~housing_pal(estimate),
              weight = 0.5,
              smoothFactor = 0.2,
              fillOpacity = 0.5,
              label = ~estimate) %>%
  addLegend(
    position = "bottomright",
    pal = housing_pal,
    values = housing_value$estimate,
    title = "Median Home Value ($) - 2019"
  )
```
This is a map of the median home values in the Bay Area counties 2019 using the ACS 1 year estimates. It shows San Francisco County ($1,233,600) with the highest median home values, then Santa Clara ($1,135,600), Marin ($1,078,800), Alameda ($882,100), Contra Costa ($687,600), Napa ($670,000), Sonoma ($664,600), and Solano ($460,500).

This is interesting to compare to the inflow and outflow maps, the regions with the most inflows and outflows such as San Francisco and San Jose are also some of the most expensive places to have homes on average.

```{r}
bay_housing_income_tract <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "tract:*",
    regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
    vars = c(
      "B06011_001E",
      "B25077_001E")
      ) %>% 
  transmute(
    tract = paste0(state, county, tract),
    house_value = B25077_001E,
    income = B06011_001E 
  ) %>% 
  filter(
    !is.na(house_value), 
    !is.na(income)
  )
    
```

```{r, progress_bar = F}
years <- 2005:2019
names(years) <- years

bay_value <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B25077_001",
    state = "CA",
    county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
    year = .x,
    survey = "acs1"
  )
}, .id = "year") %>% mutate("County" = NAME, ) %>% mutate("Median_Home_Value" = estimate)
```

```{r}
ggplot() +
  geom_point(data = bay_value, aes(x = year, y = Median_Home_Value, group = 1, color = County)) +
  theme(legend.title = element_text(face="bold"))+
  ggtitle("Estimate Median Home Values ($) for \nBay Area Counties from 2005-2019")


```
Look at the plot above to see the changes in Median Home Value across the Bay Area Counties from 2005 to 2019. 
You can see the dip in values from the 2008 housing and financial crisis, but after around 2012 median housing prices began to increase quickly and continue to increase.

```{r, progress_bar = FALSE}
bay_income1 <- get_acs(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variable = "B06011_001E",
  geometry = TRUE
) 

```

```{r}
income_pal <- colorNumeric(
  palette = "plasma",
  domain = bay_income1$estimate
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_income1,
    fillColor = ~income_pal(estimate),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~estimate
  ) %>%
  addLegend(
    position = "bottomright",
    pal = income_pal,
    values = bay_income1$estimate,
    title = "Bay Area Median Yearly Income ($) - 2019"
  )
```


```{r}
bay_income <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B06011_001",
    state = "CA",
    county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
    year = .x,
    survey = "acs1"
  )
}, .id = "year") %>% mutate("County" = NAME, ) %>% mutate("Median_Income" = estimate)
```


```{r}
ggplot() +
  geom_point(data = bay_income, aes(x = year, y = Median_Income, group = 1, color = County)) +
  theme(legend.title = element_text(face="bold"))+
  ggtitle("Estimate Median Annual Income ($) for \nBay Area Counties from 2005-2019")
```
If we look at the increase in median home prices according to the ACS 1 year survey, we can see that housing prices have increased with median income levels for each county. 
The years after the 2008 Housing and Financial Crisis marked a significant dip in both sets of data, but after 2012 both values began to rise and kept rising. 


## Bay Area Racial Demographics by County
To better understand and work to address housing and income inequality it is critical to address and unpack racial disparities and inequality. Racial disparities are a painful and ever-present reality as a result of systemic and historic inequities and discrimination that are perpetuated by policies of the past and present.

First lets take a look at Bay Area racial demographics by county using the census data. 

```{r}
groups <- unique(bay_race$variable)

bay_points <- map_dfr(groups, ~{
  bay_race %>%
    filter(variable == .x) %>%
    st_transform(26910) %>%
    mutate(est100 = as.integer(estimate / 100)) %>%
    st_sample(size = .$est100, exact = TRUE) %>%
    st_sf() %>%
    mutate(group = .x)
}) %>%
  slice_sample(prop = 1)
```

Dot Density Map of Concentration of Different Racial Groups in the Bay Area - 2019 (ACS)
```{r}
background_tracts <- filter(bay_race, variable == "White")

tm_shape(background_tracts, 
         projection = sf::st_crs(26910)) + 
  tm_polygons(col = "white", 
              border.col = "black",
              size = 0.100,
              ) + 
  tm_shape(bay_points) +
  tm_dots(col = "group", 
          palette = "RdYlBu",
          size = 0.010, 
          title = "Census Race Groups")
```

##Racial Equity Analysis 
Equity analyses are important to conduct in order for policy makers, researchers, and community organizers to highlight communities that are under-served and allocate policies, funding, or advocacy to help address these inequities. 
```{r}
acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )
```

```{r}
census_race_categories <- 
  c(
    "White Alone",
    "Black or African American",
    "American Indian and Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races"
  )

bay_race_tracts <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "tract:*",
      regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
      vars = paste0("B19001",LETTERS[x],"_001E")
    ) %>%
      mutate(
        tract = paste0(state, county, tract) %>% as.numeric(),
        race = census_race_categories[x]
      ) %>% 
      select(
        tract,
        race,
        estimate = paste0("B19001",LETTERS[x],"_001E")
      )
  })
```


```{r}
bay_mobility_race <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = paste0("group(B07004", LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,"mobility"),
        sep = "!!"
      ) %>% 
      filter(!is.na(mobility)) %>% 
      mutate(race = census_race_categories[x])
  })
```

```{r}
bay_income_race_total <-
  bay_income_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate, na.rm = T)) %>% 
  mutate(income = "Total")
```

```{r}
bay_mobility_race_total <-
  bay_mobility_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate, na.rm = T)) %>% 
  mutate(mobility = "Total")
```

```{r}
bay_income_race %>% 
  group_by(income, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_income_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = income %>% factor(levels = rev(c("Total",unique(bay_income_race$income)))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(bay_income_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Household income",
    y = "Proportion of households",
    title = "Bay Area household income by race",
    fill = "Race of householder"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```

This plot shows the proportions of Bay Area Household Income by Race with the American Communities Survey (ACS) 5 year data. This plot shows that "White Alone" households have the highest proportion of income across all income brackets but especially for the total, then "Asian Alone" and the lowest proportions are "American Indian and Alaska Native Alone" and "Native Hawaiian and Other Pacific Islander Alone."  

```{r}

bay_mobility_race %>% 
  group_by(mobility, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_mobility_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = mobility %>% factor(levels = rev(c("Total",unique(bay_mobility_race$mobility)))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(bay_mobility_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Household Mobility",
    y = "Proportion of households",
    title = "Bay Area household mobility by race",
    fill = "Race of householder"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```
This plot shows the proportions of Bay Area Household Mobility by Race with the American Communities Survey (ACS) 5 year data. The "mobility" is determined by the individual income for a current residence. ???
"White Alone" households have the highest proportion of geographic mobility in all categories except "moved from abroad" in which "Asian Alone" households have the highest proportion. "American Indian or Alaska Native Alone" and "Native Hawaiian and Other Pacific Islander Alone" have the lowest proportion of geographic mobility for every category (besides "Some Other Race Alone"). 

The geographic mobility proportions are similar to the ordering of household income by race. While both plots show significant disparities, interestingly, there is more equity or equal distribution in the geographic mobility analysis. 

## Historic Redlining in the Bay Area

As a trade-off for the New Deal policies and social programs to address the Great Depression in the U.S., the Home Owners' Loan Corporation (HOLC) was allowed to make maps that "graded" regions into four categories ('A', 'B', 'C', and 'D'). These grading schemes were on a large part build on the racial and economic demographics of the regions. The red markers, for 'high risk,' is where we get the term "redlining."
These redlining boundaries and categroizations for which neighborhoods were deemed 'high-risk' for loans or other programs has deeply impacted the Bay Area and the broader U.S. 

```{r}
bay_tracts <-
    get_acs(
      geography="tract", 
      variables = "B06011_001", 
      state="CA", 
      county=c("Alameda","Contra Costa", "Marin", "Napa", 
               "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
      year=2019, 
      survey='acs5',
      geometry=TRUE)
```

```{r}
variables <- load_variables(2019, 'acs5', cache = TRUE)
```

```{r}
bay_tracts_processed <- bay_tracts %>%
  rename_all(tolower) %>%
  st_transform(26910) %>%
  mutate(tot_area = st_area(geometry))
```

```{r}
holc <- st_read('holc.geojson')
```

```{r}
holc_proc <- holc %>%
  filter(state=="CA") %>%
  st_transform(26910)
```

```{r}
holc_fixed <- st_buffer(holc_proc, 0.0)
int <- st_intersection(holc_fixed, bay_tracts_processed)

view(int)
```

```{r}
int_area <- int %>%
  mutate(holc_area = st_area(geometry)) %>%
  st_drop_geometry() %>%
  group_by(geoid, holc_grade) %>%
  summarize(
    holc_area = as.numeric(sum(holc_area)),
    tot_area = as.numeric(max(tot_area))
  ) 
#aligning geometries of holc data + census tracts
```

```{r}
holc_areas <- merge(
    bay_tracts, 
    int_area, 
    by.x = "GEOID", 
    by.y = "geoid",
    all.x = TRUE
  ) %>% 
  filter(!is.na(holc_area))

view(holc_areas)
```
Map of HOLC Bay Area Regions with Corresponding Grades - 1937
```{r}
mapview(
  holc_areas,
  zcol = "holc_grade",
  col.regions = colorRamps::green2red
  )

```
This map shows .... by census tract --- (merged with ACS 2019 census tract geometries)...

HOLC Grading Scheme:
Grade A "Best" 
Grade B "Still Desirable"
Grade C "Definitely Declining"
Grade D "Hazardous" 

These grades correspond to how lenders interact with residents and by consequence have perpetuated cycles of discrimination and inequality in the Bay Area and beyond. 

HOLC only assessed San Francisco County, the "Oakland Area" (parts of Alameda and Contra Costa Counties, and the "San José Area" (part of Santa Clara County).

When looking at the maps of the HOLC grades from 1937, we can see many similarities to the contemporary racial and demographical information. About 80 years later, and the policy choices of late 1930s' legacy is very much a reality that impact the quality of life and discrimination faced by many in the Bay Area.  

Especially when considering the higher concentration of white households in these "A" grade neighborhoods still today being correlated with high concentrations of wealth, we can still see the impacts of these institutionalized forms of discrimination today.

## Historic HOLC Gradings Compared with Contemporary Racial Demographics
```{r}
bay_race <- get_acs(
  geography = "tract",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variables = c(White = "B03002_003",
                Black = "B03002_004",
                Native_American = "B03002_005",
                Asian = "B03002_006",
                Hispanic_or_Latino = "B03002_012"),
  summary_var = "B02001_001",
  geometry = TRUE,
  key = "ffe5c482aad3c87fd918ba8a6bd5967ae4228456"
) 


bay_race_percent <- bay_race %>%
  mutate(percent = 100 * (estimate / summary_est))
#bay racial demographics + percentages
```

```{r}
bay_asian <- filter(bay_race_percent, variable == "Asian")%>% filter(!is.na(percent))
bay_black <- filter(bay_race_percent, variable == "Black")%>% filter(!is.na(percent))
bay_hisp <- filter(bay_race_percent, variable == "Hispanic_or_Latino")%>% filter(!is.na(percent))
bay_nat_am <- filter(bay_race_percent, variable == "Native_American")%>% filter(!is.na(percent))
bay_white <- filter(bay_race_percent, variable == "White")%>% filter(!is.na(percent))
```

These maps show the percentages of people identified in the census race categories as "Asian," "Black," "Hispanic or Latino," "Native American," or "White" living in the different Bay Area census tracts. While the census also tracks other racial and ethnic categories, for the sake of the HOLC comparative analysis, these bigger categories will serve as the source of analysis and comparison to the HOLC gradings of different census tracts from 1937. 

```{r}
bay_asian_pal <- colorNumeric(
  palette = "Blues",
  domain = bay_asian$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_asian,
    fillColor = ~bay_asian_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 2,
    label = ~percent
  ) %>% addPolygons(
    data = holc_areas,
    color = none,
    weight = 0.25,
    fillOpacity = 0,
    label = ~holc_grade
  ) %>%
  addLegend(
    position = "bottomright",
    pal = bay_asian_pal,
    values = bay_asian$percent,
    title = "Percentage of Asian <br> Identifying by County - 2019 <br> with Overlayed HOLC Grades"
  )
```


```{r}
bay_black_pal <- colorNumeric(
  palette = "Blues",
  domain = bay_black$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_black,
    fillColor = ~bay_black_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 2,
    label = ~percent
  ) %>% addPolygons(
    data = holc_areas,
    color = none,
    weight = 0.25,
    fillOpacity = 0,
    label = ~holc_grade
  ) %>%
  addLegend(
    position = "bottomright",
    pal = bay_black_pal,
    values = bay_black$percent,
    title = "Percentage of Black <br> Identifying by County - 2019 <br> with Overlayed HOLC Grades"
  )
```


```{r}
bay_hisp_pal <- colorNumeric(
  palette = "Blues",
  domain = bay_hisp$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_hisp,
    fillColor = ~bay_hisp_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 2,
    label = ~percent
  )  %>% addPolygons(
    data = holc_areas,
    color = none,
    weight = 0.25,
    fillOpacity = 0,
    label = ~holc_grade
  ) %>%
  addLegend(
    position = "bottomright",
    pal = bay_hisp_pal,
    values = bay_hisp$percent,
    title = "Percentage of Hispanic or Lation <br> Identifying by County - 2019<br> with Overlayed HOLC Grades"
  )
```

```{r}
bay_nat_am_pal <- colorNumeric(
  palette = "Blues",
  domain = bay_nat_am$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_nat_am,
    fillColor = ~bay_nat_am_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 2,
    label = ~percent
  )  %>% addPolygons(
    data = holc_areas,
    color = none,
    weight = 0.25,
    fillOpacity = 0,
    label = ~holc_grade
  ) %>%
  addLegend(
    position = "bottomright",
    pal = bay_nat_am_pal,
    values = bay_nat_am$percent,
    title = "Percentage of Native American <br> Identifying by County - 2019 <br> with Overlayed HOLC Grades"
  )
```

```{r}
bay_white_pal <- colorNumeric(
  palette = "Blues",
  domain = bay_white$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_white,
    fillColor = ~bay_white_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 2,
    label = ~percent
  ) %>% addPolygons(
    data = holc_areas,
    color = none,
    weight = 0.25,
    fillOpacity = 0,
    label = ~holc_grade
  ) %>% addLegend(
    position = "bottomright",
    pal = bay_white_pal,
    values = bay_white$percent,
    title = "Percentage of White <br> Identifying by County - 2019 <br> with Overlayed HOLC Grades"
  )
```

```{r}
pums_2019_1yr <- getCensus(
  name = "acs/acs1/pums",
  vintage = 2019,
  region = "public use microdata area:*", 
  regionin = "state:06",
  vars = c(
    "SERIALNO",
    "SPORDER",
    "PWGTP",
    "WGTP",
    "NP",
    "HHL",
    "HINCP",
    "TEN"
  )
)
```


```{r}

ca_pumas <-
  pumas("CA", cb = T, progress_bar = F)

bay_county_names <-
  c(
    "Alameda",
    "Contra Costa",
    "Marin",
    "Napa",
    "San Francisco",
    "San Mateo",
    "Santa Clara",
    "Solano",
    "Sonoma"
  )

bay_counties <-
  counties("CA", cb = T, progress_bar = F) %>%
  filter(NAME %in% bay_county_names)

bay_pumas <-
  ca_pumas %>% 
  st_centroid() %>% 
  .[bay_counties, ] %>% 
  st_drop_geometry() %>% 
  left_join(ca_pumas %>% select(GEOID10)) %>% 
  st_as_sf()

bay_pums <-
  pums_2019_1yr %>% 
  mutate(
    PUMA = str_pad(public_use_microdata_area,5,"left","0")
  ) %>% 
  filter(PUMA %in% bay_pumas$PUMACE10)
```

```{r}
bay_pums_lowinc <-
  bay_pums %>% 
  filter(!duplicated(SERIALNO)) %>% 
  mutate(
    WGTP = as.numeric(WGTP),
    lowinc = ifelse(
        (HINCP < 100000),
      WGTP,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    perc_lowinc =
      sum(lowinc, na.rm =T)/sum(WGTP, na.rm = T)*100
  ) %>% 
  left_join(
    bay_pumas %>% 
      select(PUMACE10),
    by = c("PUMA" = "PUMACE10")
  ) %>% 
  st_as_sf()
```


## Map of Percent of Households in each PUMA (census microdata) that have an annual income of less than 100K.
This map shows the income percentage data as in the previous map as but with the HOLC 1937 gradings overlayed on the regions that were included in those redlining policies. Double tap on different regions to zoom-in and compare the gradings that were highly based on race and socioeconomic status in the late 1930s to the current day socio-economic breakdowns of the Bay Area.
```{r}
pums_pal <- colorNumeric(
  palette = "Blues",
  domain = bay_pums_lowinc$perc_lowinc
)

lowinc_map <- leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = bay_pums_lowinc,
    fillColor = ~pums_pal(perc_lowinc),
    color = "white",
    opacity = 0.5,
    fillOpacity = 2,
    weight = 1,
    label = ~paste0(
      round(perc_lowinc), 
      "% of households making less than $100K"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = bay_pums_lowinc,
    pal = pums_pal,
    values = ~perc_lowinc,
    title = "% Households making <br> less than $100K"
  )

lowinc_map
```

For many Bay Counties, such as San Francisco, households making less than 100K annually are cosnidered low income by HUD (U.S. Dept. of Housing and Urban Development). While this varies county to county, the Bay Area across all counties is quite an expensive place to live and making less than 100K annually can be considered lower income for this comparative analysis.  

## Map of Percent of Households with annual income of less than 100K (2019) with historic HOLC Gradings.
```{r}
lowinc_map %>% addPolygons(
    data = (holc_areas),
    color = "white",
    fillColor = FALSE,
    fillOpacity = 0,
    weight = 0.2,
    opacity = 0.2,
    label = ~holc_grade
  ) 
                            
```




