---
title: "Assessing Bay Area Housing and Income Equity and Demographics"
author: "Bella Raja"
output: html_document
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = F, warning= F, message=F, progress_bar =F)
```

```{r}
library(tidyverse)
library(censusapi)
library(tigris)
library(leaflet)
library(mapview)
library(sf)
library(tidycensus)
library(plotly)
library(dplyr)
library(tmap)
library(mapdeck)



Sys.setenv(CENSUS_KEY = "ffe5c482aad3c87fd918ba8a6bd5967ae4228456")
Sys.getenv("ffe5c482aad3c87fd918ba8a6bd5967ae4228456")
```



In this project, I wanted to analyze migration patterns in the Bay Area which led me to questions of housing and hosuing prices as related to income concentration. 
I also wanted to assess demographics in the Bay Area and examine different equity analyses and indices of segregation.
Finally, I related all of these contemporary findings to the legacy of historic redlining and economic risk "grades" given to different Bay Area regions. 



## Bay Area Migration - 2019


```{r}
acs_vars_2019_5yr <-
  listCensusMetadata(
    name = "2019/acs/acs5",
    type = "variables"
  )
```


```{r, progress_bar = F }
acs_vars_2019_1yr <-
  listCensusMetadata(
    name = "2019/acs/acs1",
    type = "variables"
  )

bay_mobility_current_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07010)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)
```

```{r}
bay_mobility_current_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07010)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"mobility","temp","income"),
    sep = "!!"
  ) %>% 
  mutate(
    income = ifelse(
      temp == "No income",
      temp,
      income
    ),
    mobility = ifelse(
      mobility %in% c("Same house 1 year ago:", "Moved within same county:"),
      "Here since last year",
      "Inflow"
    )
  ) %>% 
  filter(!is.na(income)) %>% 
  group_by(mobility, income) %>% 
  summarize(estimate = sum(estimate))

#Mobility in Bay Area
```

```{r}
bay_mobility_lastyear_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07410)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable)
#Inflow since last year
```

```{r}
bay_mobility_lastyear_19 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2019,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07410)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"mobility","temp","income"),
    sep = "!!"
  ) %>% 
  mutate(
    income = ifelse(
      temp == "No income",
      temp,
      income
    ),
    mobility = ifelse(
      mobility %in% c("Same house:", "Moved within same county:"),
      "Here since last year",
      "Outflow"
    )
  ) %>% 
  filter(!is.na(income)) %>% 
  group_by(mobility, income) %>% 
  summarize(estimate = sum(estimate))

view(bay_mobility_lastyear_19)
#outflow
```
 



```{r}
bay_mobility_current_18 <- 
  getCensus(
    name = "acs/acs1",
    vintage = 2018,
    region = "county:001,013,041,055,075,081,085,095,097",
    regionin = "state:06",
    vars = c("group(B07010)")
  ) %>% 
  select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
  pivot_longer(
    ends_with("E"),
    names_to = "variable",
    values_to = "estimate"
  ) %>%
  left_join(
    acs_vars_2019_1yr %>% 
      select(name, label), 
    by = c("variable" = "name")
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA, NA, "mobility","temp","income"),
    sep = "!!"
  ) %>% 
  mutate(
    income = ifelse(
      temp == "No income",
      temp,
      income
    ),
    mobility = "Here last year"
  ) %>% 
  filter(!is.na(income)) %>% 
  group_by(mobility, income) %>% 
  summarize(estimate = sum(estimate))
#total pop 2018
```


```{r}
bay_flows_19 <-
  rbind(
    bay_mobility_current_18,
    bay_mobility_lastyear_19 %>% 
      filter(mobility == "Outflow"),
    bay_mobility_current_19 %>% 
      filter(mobility == "Inflow"),
    bay_mobility_current_19 %>% 
      group_by(income) %>% 
      summarize(estimate = sum(estimate)) %>% 
      mutate(mobility = "Here this year")
  ) %>% 
  pivot_wider(
    names_from = mobility,
    values_from = estimate
  ) %>% 
  mutate(
    `External net` = Inflow - Outflow,
    `Internal net` = `Here this year` - `Here last year` - `External net`,
  ) %>% 
  select(
    `Income tier` = income, 
    `Internal net`,
    `External net`,
    `Here last year`, 
    `Here this year`, 
    Outflow, 
    Inflow
  )


bay_flows_19
```




```{r}
token <- "pk.eyJ1IjoiYmVsbGFyIiwiYSI6ImNrd3JjNmFkOTB2MmQycWp2Mm42OXc5dXgifQ.pttZ3IYhuP9cq19jSqvIzA"

bay_inflow <- get_flows(
  geography = "county",
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  geometry = TRUE,
  year = 2019
) %>%
  filter(variable == "MOVEDIN") %>%
  na.omit() %>%
  arrange(desc(estimate))

# finding inflows again but this time using tidycensus flows function
```

The arcs show where people have moved from into the Bay Area in 2019. 



```{r}
bay_outflow <- get_flows(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  geometry = TRUE,
  key = "ffe5c482aad3c87fd918ba8a6bd5967ae4228456"
) %>%
  filter(variable == "MOVEDOUT") %>%
  na.omit() %>%
  arrange(desc(estimate))
view(bay_outflow)
```
These 3D Maps show the inflow and outflow for the Bay Area. 
I am showing the top 50 regions that are the inflow and outflow origins and destinations
To navigate these maps....

Bay Area Inflow Regions - 2019                     
```{r}
bay_inflow  %>%
  slice_max(estimate, n = 50) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = token) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight")
```
This map shows the ...


Bay Area Outflow Regions - 2019
```{r}
bay_outflow  %>%
  slice_max(estimate, n = 50) %>%
  mutate(weight = estimate / 500) %>%
  mapdeck(token = token) %>%
  add_arc(origin = "centroid2",
          destination = "centroid1",
          stroke_width = "weight")


```
The arcs show where people have moved to and from the Bay Area in 2019. 

## Housing Costs in the Bay Area
```{r}
housing_value <- get_acs(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variables = "B25077_001",
  survey = "acs1",
  geometry = TRUE,
  resolution = "20m"
)
```

```{r}
housing_pal <- colorNumeric(
  palette = "plasma",
  domain = housing_value$estimate
)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(data = housing_value,
              color = ~housing_pal(estimate),
              weight = 0.5,
              smoothFactor = 0.2,
              fillOpacity = 0.5,
              label = ~estimate) %>%
  addLegend(
    position = "bottomright",
    pal = housing_pal,
    values = housing_value$estimate,
    title = "Median Home Value ($)"
  )
```
This is a map of the median home values in the Bay Area counties 2019. It shows San Francisco County ($1,233,600) with the highest median home values, then Santa Clara ($1,135,600), Marin ($1,078,800), Alameda ($882,100), Contra Costa ($687,600), Napa ($670,000), Sonoma ($664,600), and Solano ($460,500).

This is interesting to compare to the inflow and outflow maps, the regions with the most inflows and outflows such as San Francisco and San Jose are also some of the most expensive places to have homes on average.

## Assesing Bay Area Income and Housing

```{r}
census_race_categories <- 
  c(
    "White Alone",
    "Black or African American",
    "American Indian and Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races"
  )

bay_race_tracts <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "tract:*",
      regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
      vars = paste0("B19001",LETTERS[x],"_001E")
    ) %>%
      mutate(
        tract = paste0(state, county, tract) %>% as.numeric(),
        race = census_race_categories[x]
      ) %>% 
      select(
        tract,
        race,
        estimate = paste0("B19001",LETTERS[x],"_001E")
      )
  })
```

```{r}
bay_income_race <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = paste0("group(B19001",LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,"income"),
        sep = "!!"
      ) %>% 
      filter(!is.na(income)) %>% 
      mutate(race = census_race_categories[x])
  })
view(bay_income_race)
```

```{r progress_bar = F}


bay_income1 <- get_acs(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variable = "B06011_001E",
  geometry = TRUE
) 

```

```{r}
income_pal <- colorNumeric(
  palette = "plasma",
  domain = bay_income1$estimate
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_income1,
    fillColor = ~income_pal(estimate),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~estimate
  ) %>%
  addLegend(
    position = "bottomright",
    pal = income_pal,
    values = bay_income1$estimate,
    title = "Bay Area Median Yearly Income ($)"
  )
```
The median yearly income by county also corresponds similarly with the median housing prices by county. 

```{r}
bay_housing_income_tract <- 
  getCensus(
    name = "acs/acs5",
    vintage = 2019,
    region = "tract:*",
    regionin = "state:06+county:001,013,041,055,075,081,085,095,097",
    vars = c(
      "B06011_001E",
      "B25077_001E")
      ) %>% 
  transmute(
    tract = paste0(state, county, tract),
    house_value = B25077_001E,
    income = B06011_001E 
  ) %>% 
  filter(
    !is.na(house_value), 
    !is.na(income)
  )
    
view(bay_housing_income_tract)
```

```{r, progress_bar = F}
years <- 2005:2019
names(years) <- years

bay_value <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B25077_001",
    state = "CA",
    county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
    year = .x,
    survey = "acs1"
  )
}, .id = "year") %>% mutate("County" = NAME, ) %>% mutate("Median_Home_Value" = estimate)
```

```{r}
ggplot() +
  geom_point(data = bay_value, aes(x = year, y = Median_Home_Value, group = 1, color = County)) +
  theme(legend.title = element_text(face="bold"))+
  ggtitle("Estimate Median Home Values ($) for \nBay Area Counties from 2005-2019")


```

```{r}
bay_income <- map_dfr(years, ~{
  get_acs(
    geography = "county",
    variables = "B06011_001",
    state = "CA",
    county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
    year = .x,
    survey = "acs1"
  )
}, .id = "year") %>% mutate("County" = NAME, ) %>% mutate("Median_Income" = estimate)
```


```{r}
ggplot() +
  geom_point(data = bay_income, aes(x = year, y = Median_Income, group = 1, color = County)) +
  theme(legend.title = element_text(face="bold"))+
  ggtitle("Estimate Median Annual Income ($) for \nBay Area Counties from 2005-2019")
```
If we look at the increase in median home prices according to the ACS 1 year survey, we can see that housing prices have increased with median income levels for each county. 
The years after the 2008 Housing and Financial Crisis marked a significant dip in both sets of data, but after 2012 both values began to rise and kept rising. 

##Racial Equity Analysis 
Racial disparities are a painful reality as a result of systemic and historic inequities and discrimination. 
Equity analyses are important to conduct in order for policy makers, researchers, and community organizers to highlight communities that are under-served and allocate policies, funding, or advocacy to help address these inequities. 

```{r}
bay_mobility_race <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = paste0("group(B07004", LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,"mobility"),
        sep = "!!"
      ) %>% 
      filter(!is.na(mobility)) %>% 
      mutate(race = census_race_categories[x])
  })
```


```{r}
bay_income_race_total <-
  bay_income_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate, na.rm = T)) %>% 
  mutate(income = "Total")
```

```{r}
bay_mobility_race_total <-
  bay_mobility_race %>% 
  group_by(race) %>% 
  summarize(estimate = sum(estimate, na.rm = T)) %>% 
  mutate(mobility = "Total")
```

```{r}
bay_income_race %>% 
  group_by(income, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_income_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = income %>% factor(levels = rev(c("Total",unique(bay_income_race$income)))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(bay_income_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Household income",
    y = "Proportion of households",
    title = "Bay Area household income by race",
    fill = "Race of householder"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```



This plot shows the proportions of Bay Area Household Income by Race with the American Communities Survey (ACS) 5 year data. This plot shows that "White alone households" have the highest proportion of income, then "Asian Alone" and the lowest proportions are "American Indian and Alaska Native Alone" and "Native Hawaiian and Other Pacific Islander Alone."  

```{r}

bay_mobility_race %>% 
  group_by(mobility, race) %>% 
  summarize(estimate = sum(estimate)) %>% 
  rbind(bay_mobility_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = mobility %>% factor(levels = rev(c("Total",unique(bay_mobility_race$mobility)))),
      y = estimate,
      fill = race %>% factor(levels = rev(unique(bay_mobility_race$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Household Mobility",
    y = "Proportion of households",
    title = "Bay Area household mobility by race",
    fill = "Race of householder"
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )  +
  guides(
    fill = guide_legend(
      reverse = T
    )
  )
```
This plot shows the proportions of Bay Area Household Mobility by Race with the American Communities Survey (ACS) 5 year data. The "mobility" is determined by the individual income for a current residence. ???


## Bay Area Racial Demographics by Census County
```{r}


bay_race <- get_acs(
  geography = "county",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variables = c(White = "B03002_003",
                Black = "B03002_004",
                Native = "B03002_005",
                Asian = "B03002_006",
                Hispanic = "B03002_012"),
  summary_var = "B03002_001",
  geometry = TRUE
) 



bay_race_percent <- bay_race %>%
  mutate(percent = 100 * (estimate / summary_est))

```

```{r}
groups <- unique(bay_race$variable)

bay_points <- map_dfr(groups, ~{
  bay_race %>%
    filter(variable == .x) %>%
    st_transform(26910) %>%
    mutate(est100 = as.integer(estimate / 100)) %>%
    st_sample(size = .$est100, exact = TRUE) %>%
    st_sf() %>%
    mutate(group = .x)
}) %>%
  slice_sample(prop = 1)
```

Dot Density Map of Concentration of Different Racial Groups in the Bay Area - 2019 (ACS)
```{r}
background_tracts <- filter(bay_race, variable == "White")

tm_shape(background_tracts, 
         projection = sf::st_crs(26910)) + 
  tm_polygons(col = "white", 
              border.col = "black",
              size = 0.100,
              ) + 
  tm_shape(bay_points) +
  tm_dots(col = "group", 
          palette = "RdYlBu",
          size = 0.010, 
          title = "Census Race Groups")
```


## Historic Redlining in the Bay Area

As a trade-off for the New Deal policies and social programs to address the Great Depression in the U.S., the Home Owners' Loan Corporation (HOLC) was allowed to make maps that "graded" regions into four categories ('A', 'B', 'C', and 'D'). These grading schemes were on a large part build on the racial and economic demographics of the regions. The red markers, for 'high risk,' is where we get the term "redlining."
These redlining boundaries and categroizations for which neighborhoods were deemed 'high-risk' for loans or other programs has deeply impacted the Bay Area and the broader U.S. 

```{r}
bay_tracts <-
    get_acs(
      geography="tract", 
      variables = "B06011_001", 
      state="CA", 
      county=c("Alameda","Contra Costa", "Marin", "Napa", 
               "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
      year=2019, 
      survey='acs5',
      geometry=TRUE)
```

```{r}
variables <- load_variables(2019, 'acs5', cache = TRUE)
```

```{r}
bay_tracts_processed <- bay_tracts %>%
  rename_all(tolower) %>%
  st_transform(26910) %>%
  mutate(tot_area = st_area(geometry))
```

```{r}
holc <- st_read('holc.geojson')
```

```{r}
holc_proc <- holc %>%
  filter(state=="CA") %>%
  st_transform(26910)
```

```{r}
holc_fixed <- st_buffer(holc_proc, 0.0)
int <- st_intersection(holc_fixed, bay_tracts_processed)

view(int)
```

```{r}
int_area <- int %>%
  mutate(holc_area = st_area(geometry)) %>%
  st_drop_geometry() %>%
  group_by(geoid, holc_grade) %>%
  summarize(
    holc_area = as.numeric(sum(holc_area)),
    tot_area = as.numeric(max(tot_area))
  ) 
#aligning geometries of holc data + census tracts
```

```{r}
holc_areas <- merge(
    bay_tracts, 
    int_area, 
    by.x = "GEOID", 
    by.y = "geoid",
    all.x = TRUE
  ) %>% 
  filter(!is.na(holc_area))

view(holc_areas)
```
Map of HOLC Bay Area Regions with Corresponding Grades - 1937
```{r}
mapview(
  holc_areas,
  zcol = "holc_grade",
  col.regions = colorRamps::green2red
  )

```
This map shows .... by census tract --- (merged with ACS 2019 census tract geometries)...

HOLC Grading Scheme:
Grade A "Best" 
Grade B "Still Desirable"
Grade C "Definitely Declining"
Grade D "Hazardous" 

These grades correspond to how lenders interact with residents and by consequence have perpetuated cycles of discrimination and inequality in the Bay Area and beyond. 

HOLC only assessed San Francisco County, the "Oakland Area" (parts of Alameda and Contra Costa Counties, and the "San José Area" (part of Santa Clara County).

When looking at the maps of the HOLC grades from 1937, we can see many similarities to the contemporary racial and demographical information. About 80 years later, and the policy choices of late 1930s' legacy is very much a reality that impact the quality of life and discrimination faced by many in the Bay Area.  

Especially when considering the higher concentration of white households in these "A" grade neighborhoods still today being correlated with high concentrations of wealth, we can still see the impacts of these institutionalized forms of discrimination today.

### Bay Area Racial Demographics
```{r}


bay_race <- get_acs(
  geography = "tract",
  year = 2019,
  state = "CA",
  county = c("Alameda","Contra Costa", "Marin", "Napa", "San Francisco", "San Mateo", "Santa Clara", "Solano", "Sonoma"),
  variables = c(White = "B03002_003",
                Black = "B03002_004",
                Native = "B03002_005",
                Asian = "B03002_006",
                Hispanic = "B03002_012"),
  summary_var = "B03002_001",
  geometry = TRUE,
  key = "ffe5c482aad3c87fd918ba8a6bd5967ae4228456"
) 


bay_race_percent <- bay_race %>%
  mutate(percent = 100 * (estimate / summary_est))



view(bay_race)
```

```{r}
bay_asian <- filter(bay_race_percent, variable == "Asian")
bay_black <- filter(bay_race_percent, variable == "Black")
bay_hisp <- filter(bay_race_percent, variable == "Hispanic")
bay_nat_am <- filter(bay_race_percent, variable == "Native")
bay_white <- filter(bay_race_percent, variable == "White")
```


```{r}
bay_asian_pal <- colorNumeric(
  palette = "Reds",
  domain = bay_asian$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_asian,
    fillColor = ~bay_asian_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~percent
  ) %>% addPolygons(
    data = holc_areas,
    color = "white",
  label = ~holc_grade
  )
```

```{r}
bay_black_pal <- colorNumeric(
  palette = "Reds",
  domain = bay_black$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_black,
    fillColor = ~bay_black_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~percent
  )
```


```{r}
bay_hisp_pal <- colorNumeric(
  palette = "Reds",
  domain = bay_hisp$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_hisp,
    fillColor = ~bay_hisp_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~percent
  )
```



```{r}
bay_nat_am_pal <- colorNumeric(
  palette = "Reds",
  domain = bay_nat_am$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_nat_am,
    fillColor = ~bay_nat_am_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~percent
  )
```


```{r}
bay_white_pal <- colorNumeric(
  palette = "Reds",
  domain = bay_white$percent
    
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = bay_white,
    fillColor = ~bay_white_pal(percent),
    color = "white",
    weight = 0.5,
    fillOpacity = 0.5,
    label = ~percent
  )
```



???
```{r}
acs_vars_1937_5yr <-
  listCensusMetadata(
    name = "2014/acs/acs5",
    type = "variables"
  )
```

```{r}
bay_income_race_193 <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
      region = "county:001,013,041,055,075,081,085,095,097",
      regionin = "state:06",
      vars = paste0("group(B19001",LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "estimate"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,"income"),
        sep = "!!"
      ) %>% 
      filter(!is.na(income)) %>% 
      mutate(race = census_race_categories[x])
  })
```



